---
title: "hyfo easy start"
author: "Yuanchao Xu"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{hyfo easy start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
`hyfo` is designed for hydrology and forecasting anaylasis, containing a number of tools including data extration, data processing and data visulization. There are two main functions in the package:

1. Hydrology
* Providing tools from raw data extration to model input data.
* e.g., data extraction from file, precipitation gap filler, annual precipitation calculation.

2. Forecasting
* Providing tools from forecasting data visulization and analysis.
* e.g., get spatial maps, data analysis and bias analysis.

## Vignette Info

Note that the functions end with `_anarbe` are the functions designed specially for some case in Spain, those
functions may not be applied to all the cases.

## Hydrology

### collectData



The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
