---
title: "hyfo Easy Start"
author: "Yuanchao Xu"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 3
  html_vignette:
    toc: true
    toc_depth: 3
  pdf_vignette:
    toc: true
    toc_depth: 3
    
vignette: >
  %\VignetteIndexEntry{hyfo easy start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{ASCII}
---
# hyfo 

#### is designed for hydrology and forecasting anaylasis, containing a number of tools including data extration, data processing and data visulization. There are two main parts in the package, as well as in this mannual:

1. Hydrology
* Providing tools from raw data extration to final precipitation data used by model.
* e.g., data extraction from file, precipitation gap filler, annual precipitation calculation.

2. Forecasting
* Providing tools from forecasting data visulization and analysis.
* e.g., get spatial maps, data analysis and bias analysis.

##### Note
* For the hydrology tools part, the minimum time unit is a day, i.e., it mainly focuses on water resource and some long term analysis. For flood analysis part, it will be added in future.

* For the forecasting tools part, `hyfo` mainly focuses on the post processing of the gridData derived from forecasts or other sources. The input is a list file, usually a result from  `loadGridData{ecomsUDG.Raccess}` or `loadECOMS{ecomsUDG.Raccess}`. Pacakage[`{ecomsUDG.Raccess}`](https://github.com/SantanderMetGroup/ecomsUDG.Raccess)is designed for getting access to different dataset, and also can load grid file (like netcdf file) directly.

* The functions end with `_anarbe` are the functions designed specially for some case in Spain, those functions mostly are about data collection of the anarbe catchment, which will be introduced in the end fo this mannual.

##### Installation

You can go [here](http://yuanchao-xu.github.io/hyfo/) to download installation file, and use IDE like Rstudio to install from file, both `tar.gz` and `zip` formats are provided.

Also you can use the following code to install the latest version.
```{}
  install.packages('devtools')
  # Ignore the warning that Rtool is not installed, unless you want other function from devtools.
  # If you have "devtools" installed already, you just need to run the following code.
  devtools::install_github('Yuanchao-Xu/hyfo')
```

# 1. Hydrology

##### Note

If you are an experienced R user, and know how to read data in R, deal with dataframe, generate date and list, please start from next charpter, "1.2 Rainfall Analysis"



## 1.1 Data Preparation

### 1.1.1 From File

`hyfo` does provide a common tool for collecting data from different type of files, including "txt", 

"csv" and "excel", which has to be assigned to the argument `fileType`.

Now let's use internal data as an example.

```{r, }
library(hyfo)#load the package.
# get the folder containing different csv (or other type) files.
file <- system.file("extdata", "1999.csv", package = "hyfo")
folder <- strsplit(file, '1999')[[1]][1]

# Extract and combine content from different files and in each file, the extracted zone is 
# from row 10 to row 20, Column 1 to column2.
a <- collectData(folder, fileType = 'csv', range = c(10, 20, 1, 2))
str(a)
```

`a` cannot be directly inputed in `hyfo`, it still needs some process.
```{r}
# Check the date to see if it follows the format in ?as.Date(), if not, 
# use as.Date to convert. 
a <- data.frame(a)
#get date
date <- a[, 1]

# The original format is d/m/year, convert to formal format.
date <- as.Date(date, format = '%d/%m/%Y')
a[, 1] <- date

# Now a has become `a` time series dataframe, which is the atom element of the analysis. 
#`hyfo` deals with list containing different time series dataframe. In this example, 
#there is only one dataframe, and more examples please refer to the following chapter.
datalist <- list(a)

# Use getAnnual as an example, here since `a` is not a complete time series, 
# the result is only base on the input.
# getAnnual gives the annual precipitation of each year, 
# and will be introduced in the next chapter.
getAnnual(datalist)
```


### 1.1.2 Mannually

Following example shows a simple way to generate dataframe with start date, end date, and the value. Here in the example, `sample()` is used to generate random values, while in real case it will be a vector containing time series values.
```{r, fig.show = 'hold', fig.height = 4, fig.width = 7}
# Generate timeseries datalist. Each data frame consists of a Date and a value.
library(hyfo)
AAA <- data.frame(
  Date = seq(as.Date('1990-10-28'), as.Date('1997-4-1'), 1), # Date column
  AAA = sample(1:10, length(seq(as.Date('1990-10-28'), # value column
                                as.Date('1997-4-1'), 1)), repl = TRUE))

BBB <- data.frame(
  Date = seq(as.Date('1993-3-28'), as.Date('1999-1-1'),1), 
  BBB = sample(1:10, length(seq(as.Date('1993-3-28'), 
                                as.Date('1999-1-1'),1)), repl = TRUE))

CCC <- data.frame(
  Date = seq(as.Date('1988-2-2'), as.Date('1996-1-1'),1), 
  CCC = sample(1:10, length(seq(as.Date('1988-2-2'), 
                                as.Date('1996-1-1'),1)), repl = TRUE)) 

datalist <- list(AAA, BBB, CCC)# dput() and dget() can be used to save and load list file.
a <- getAnnual(datalist)
```



## 1.2 Rainfall Analysis

Assuming we have three gauging stations named "AAA", "BBB", "CCC", the precipitation information can be get by the following:
```{r, fig.show='hold', fig.height=4, fig.width=7}
# testdl is a datalist provided by the package as a test. 
# It's a list containing different time series.
data(testdl)
a <- getAnnual(testdl)
```

As shown above, the annual precipitation and the number of missing values are shown in the figure. Knowing how many missing values you have is alway important when calculating the mean annual precipitation. 

Now we want to get the mean annual precipitation.
```{r, fig.show='hold', fig.height=4, fig.width=7}
a <- getAnnual(testdl, output = 'mean')
a
```

Mean annual precipitation is calculated, but as we can see in the figure before, it's not reliable, since there are a lot of missing values in AAA and CCC, especially in AAA, in 1993, there are more than 30 missing values in a year. So we have to decide which is the threshold for the valid record. the default is 355, which means in a year (355 or 365 days), if the valid records (not missing) exceeds 355, then this year is taken into consideration in the mean annual preicipitation calculation.
```{r, fig.show='hold', fig.height=3, fig.width=3}
getAnnual(testdl, output = 'mean', minRecords = 300)
getAnnual(testdl, output = 'mean', minRecords = 365)
```

If you are not satisfied with the title and x axis and y axis, you can assign them yourself.
```{r, fig.show='hold', fig.height=4, fig.width=7, results='hide'}
a <- getAnnual(testdl, output = 'mean', title = 'aaa', x = 'aaa', y = 'aaa')
```

If you want to calculate annual rainfall for a single dataframe containing one time series.
```{r, fig.show='hold', fig.height=4, fig.width=7}
a <- getAnnual_dataframe(testdl[[1]])
a
```



## 1.3 Extract Common Period from Different Time Series

Now we have the general information of the precipitation, if we want to use them in a model, we have to extract the common period of them, and use the common period precipitation to analyze.
```{r, fig.show='hold', fig.height=4, fig.width=7}
testdl_new <- extractPeriod(testdl, commonPeriod = TRUE )
str(testdl_new)
```

If we want to extract data from a certain period, we can assgin start and end date.

```{r, fig.show='hold', fig.height=4, fig.width=7}
# Extract period of the winter of 1994
testdl_new <- extractPeriod(testdl, startDate = '1994-12-01', endDate = '1995-03-01' )
str(testdl_new)
```



## 1.4 Fill Gaps (rainfall data gaps)

Although we have got the precipitation of the common period, we can still see that there are some missing values inside, which we should fill.
```{r, fig.show='hold', fig.height=4, fig.width=7}
testdl_new <- extractPeriod(testdl, commonPeriod = TRUE )
a <- getAnnual(testdl_new)
a
```

First we have to transform the datalist to dataframe, which can be done by the code below:
```{r, fig.show='hold', fig.height=4, fig.width=7}
df <- list2Dataframe(testdl_new)
head(df)
```

From above, we can see that in the gauging station "AAA", there are some missing value marked as "NA". Now we are going to fill these gaps.

The gap filling is based on the correlation and linear regression between each two gauging stations, correlation table, correlation Order and Linear Coefficients are also printed when doing the calculation. Details can be found in  `?fillGap`.
```{r, fig.show='hold', fig.height=4, fig.width=7}
df_filled <- fillGap(df)
head(df_filled)
```

Default correlation period is "daily", while sometimes the daily rainfall correlation of precipitation is not so strong, we can also select the correlation period.
```{r, fig.show='hold', fig.height=4, fig.width=7}
df_filled <- fillGap(df, corPeriod = 'monthly')
head(df_filled)
df_filled <- fillGap(df, corPeriod = 'yearly')
head(df_filled)
```



## 1.5 Seasonal and Monthly Precipitation

Sometimes we need to know not only the annual precipitation, but also the precipitation of a certain month or certain season.
```{r, fig.show='hold', fig.height=4, fig.width=7}
data(testdl)
# year and mon can be extracted from date.
TS  <- testdl[[1]]
year = as.numeric(format(TS[, 1], '%Y'))
month = as.numeric(format(TS[, 1], '%m'))
```
Get the mean spring precipitation.
```{r, fig.show='hold', fig.height=4, fig.width=7}
a <- getMeanPreci(TS[, 2], method = 'spring', yearIndex = year, monthIndex = month)
a
```
Get the series of spring precipitation, set `fullResults = TRUE`.
```{r, fig.show='hold', fig.height=4, fig.width=7}
a <- getMeanPreci(TS[, 2], method = 'spring', yearIndex = year, monthIndex = month,
                  fullResults = TRUE)
a
```
If missing value is excluded, set omitNA = TRUE.
```{r, fig.show='hold', fig.height=4, fig.width=7}
a <- getMeanPreci(TS[, 2], method = 'winter', yearIndex = year, monthIndex = month,
                  omitNA = TRUE, fullResults = TRUE)
a
```
Get special month precipitation, e.g. march.
```{r, fig.show='hold', fig.height=4, fig.width=7}
a <- getMeanPreci(TS[, 2], method = 3, yearIndex = year, monthIndex = month,
                  fullResults = TRUE)
a
```
We can also get annual precipitation, plot figure and assign title ans axis.
```{r, fig.show='hold', fig.height=4, fig.width=7}
a <- getMeanPreci(TS[, 2], method = 'annual', yearIndex = year, monthIndex = month,
                  fullResults = TRUE,omitNA = TRUE, plot = TRUE, x = 'aaa', title = 'aaa')
```

## 1.6 Get ensemble hydrological forecast from historical data


The basic forecasts are made from the historical data, to see, how the historical data act in the same situation. Using the same period from the historical data to generate an ensemble forcast. 

E.g., we have a period of data from 2000 to 2007, we assume 2004 to be the forecast year. Then, use 2004 as an example, the data in 2000, 2001, 2002, 2003, 2005, 2006, 2007 will be taken to generate an ensemble forecast of 6 members(except 2004).

Set example year, e.g., year 1994.
```{r}
data(testdl)

a <- testdl[[1]]
a1 <- getHisEnsem(a, example = c('1994-1-1', '1994-12-31'))

```

Example period can be any time, can be a year or some months.
```{r}
a2 <- getHisEnsem(a, example = c('1995-1-1', '1996-3-1'))
a3 <- getHisEnsem(a, example = c('1995-1-1', '1995-8-11'))
```

`interval` means the interval between each member. Check `?getHisEnsem` for detailed instruction. Default is 365, representing one year.
```{r}
# If interval is two years.
a2 <- getHisEnsem(a, example = c('1995-1-1', '1996-3-1'), interval = 730)
str(a2)
# If interval is three months.
a3 <- getHisEnsem(a, example = c('1995-1-1', '1995-8-11'), interval = 90)
str(a3)
# If interval is 171 days.
a4 <- getHisEnsem(a, example = c('1995-1-1', '1995-8-11'), interval = 171)
str(a4)
```

For some models, like MIKE NAM, it's necessary to run model a few days before the forecasting time, to warm up the model. In this case `buffer` is needed to generate the "warm up period".
```{r}
# If the model needs 14 days to warm up.
a2 <- getHisEnsem(a, example = c('1995-1-1', '1996-3-1'), interval = 730, buffer = 14)
str(a2)
```

From `str(a2)` we can see that the data has 14 more rows, and the start date is changed to "1994-12-18"

# 2. Forecasting

##### Note

If an ensemble forecasting data is loaded, there will be one dimension called  "member", by default, `hyfo` will calculate the mean of different members. If you want to see a special member, add `member` argument to `getSpatialMap`, e.g., `getSpatialMap(tgridData, method = 'meanAnnual', member = 3)`, `getPreciBar(tgridData, method = 'annual', member = 14)`

## 2.1 Plot Spatial Map

As described at the start of the mannual, `hyfo` is mainly in charge of the post processing of the forecast data. Input of `hyfo` should be the result from  `loadGridData{ecomsUDG.Raccess}` or `loadECOMS{ecomsUDG.Raccess}`. An example is included in the package. 

If we want to see the mean daily precipitation.
```{r, fig.show='hold', fig.height=4, fig.width=7}
data(tgridData)
a <- getSpatialMap(tgridData, method = 'meanAnnual')
```

There are several methods to be seleted in the function, details can be found by `?getSpatialMap`.

Sometimes there exists a great difference in the whole map, e.g., the following value, `c(100, 2, 2,6, 1,7)`, since the maximum value is too large, so in the plot, by normal plot scale, we can only recognize value 100 and the rest, it's hard for us to tell the difference between 2, 2.6, and 1.7 from the plot. In this situation, the value needs to be processed before plotting. Here `scale` provides a way to decide the plot scale.

`scale` passes the arguments to the `trans` argument in `ggplot2`. The most common scale is "sqrt" and "log10", which focus more on the minutiae. Default is "identity", which means no change to the plot scale.
```{r, fig.show='hold', fig.height=4, fig.width=7}
a <- getSpatialMap(tgridData, method = 'meanAnnual', scale = 'sqrt')
```

Here in our example, because the region is too small, and the differences is not so big, so it's not so obvious to tell from the plot. But if in a map, both dry region and wet region is included, that will be more obvious to see the difference between the plot scales.


Also, if you are not satisfied with the title, x axis and y axis, you can assgin yourself.
```{r, fig.show='hold', fig.height=4, fig.width=7}
a <- getSpatialMap(tgridData, method = 'meanAnnual', scale = 'sqrt', 
                   title = 'aaa', x = 'aaa', y = 'aaa')
```

## 2.2 Add Background (catchment and gauging stations)

The default background is the world map, while if you have other backgrounds like catchment shape file and station location file, you are welcome to import them as background.

### 2.2.1 Add catchment shape file

Catchment shape file needs to be processed with a very simple step. It's based on the package `rgdal`, details can be found by `?shp2cat`

```{r, fig.show='hold', fig.height=4, fig.width=7}
# Use the test file provided by hyfo
file <- system.file("extdata", "testCat.shp", package = "hyfo")
cat <- shp2cat(file)
# cat is the catchment file.
```

Then the catchment file `cat` can be inputed as background.

```{r, fig.show='hold', fig.height=4, fig.width=7}
a <- getSpatialMap(tgridData, method = 'meanAnnual', catchment = cat)
```

### 2.2.2 Add station locations

Points file needs to be read into dataframe, and special column has to be assigned, details can be found by `?getSpatialMap_mat`
```{r, fig.show='hold', fig.height=6, fig.width=7}
# Use the points file provided by hyfo
file <- system.file("extdata", "points.txt", package = "hyfo")
points <- read.table(file, header = TRUE, sep = ',' )
getSpatialMap(tgridData, method = 'winter', points = points, catchment = cat)

```

As can be seen above, the color of the points represents the elevation, the size of the points represents the value, e.g., rainfall value.


## 2.3 Precipitation Bar Plot

Bisides spatial map, bar plot can also be plotted. The value in the bar plot is spatially averaged, i.e. the value in the bar plot is the mean value over the region.

Annual precipitation.

```{r, fig.show='hold', fig.height=4, fig.width=7}
data(tgridData)
a <- getPreciBar(tgridData, method = 'annual')
```

Mean monthly precipitation over the whole period, with the ranges for each month. But not all kinds of bar plot have a plot range.

```{r, fig.show='hold', fig.height=4, fig.width=7}
a <- getPreciBar(tgridData, method = 'meanMonthly')
a <- getPreciBar(tgridData, method = 'meanMonthly', plotRange = FALSE)
```

Seasonal precipitation, and monthly precipitation can also be plotted. 

```{r, fig.show='hold', fig.height=4, fig.width=7}
a <- getPreciBar(tgridData, method = 'spring')# spring precipitation for each year
a <- getPreciBar(tgridData, method = 3) # march precipitation for each year
```

## 2.4 Analysis and Comparison

For some cases, analysis and comparison are necesssary, which are also provided by `hyfo`. 

There are three different kinds of output from `getSpatialMap` and `getPreciBar`, respectively, `output = 'data'`, `output = 'ggplot'` and `output = 'plot'`. 

`output = 'data'` is default in the function and do not need to be declare when input. It is mainly used in analyzing and replot the results.

`output = 'ggplot'` is used when combining different plots.

`output = 'plot'` is used when a layer output is needed. the output can be directly printed, and can be mannually combined by the plot arrange functions, e.g., `grid.arrange()`

##### Note:
**All the comparisons must be comparable, e.g.,**

* For `getSpatialMap_comb`, the maps to be compared should be with same size and resolution, in other words, they should be fully overlapped by each other. Check `?getSpatialMap_comb` for details.
* For `getPreciBar_comb`, the bar plots to be compared should belong to the same kind, e.g., spring and winter, January and December, and couldn't be spring and annual. Details can be found by `?getPreciBar_comb`

### 2.4.1 Spatial Map

The default "data" output provides a matrix, representing the raster information of the spatial map. 

```{r, fig.show='hold', fig.height=4, fig.width=7}
a <- getSpatialMap(tgridData, method = 'meanAnnual')
a
```

This matrix is upside down from what you can see from the plot. **DO NOT try to change this matrix.**
`hyfo` can deal with it.

```{r, fig.show='hold', fig.height=4, fig.width=7}
# For re-plot the matrix
b <- getSpatialMap_mat(a)

# Without title and x and y, also you can assign yourself.
b <- getSpatialMap_mat(a, title = 'aaa', x = 'aaa', y = '')
```

The matrix can be used to make different analysis and plot again.

##### Note
**If the matrix doesn't come from `getSpatialMap`, dimension name of longitude and latitude needs to be provided to the matrix, in order to be plotted.**
```{r, fig.show='hold', fig.height=4, fig.width=7, results='hide'}

a1 <- getSpatialMap(tgridData, method = 'mean')
a2 <- getSpatialMap(tgridData, method = 'max')

# To see the difference between mean value and maximum value.
b <- a2 - a1
getSpatialMap_mat(b, title = '', x = '', y = '')

# To make some changes to mean value.
b <- a1 * 3 -1
getSpatialMap_mat(b, title = '', x = '', y = '')

# Bias, variation and other analysis can also be processed 
# the same way. 
# Just apply the analysis to the matrix and 
# use getSpatialMap_mat to plot.
```

If multi-plot is needed, `hyfo` can also combine different plots together. Use `output = ggplot`, which gives back the a special format that can be easily used by `ggplot2`

```{r, fig.show='hide', fig.height=4, fig.width=7}

a1 <- getSpatialMap(tgridData, method = 'spring', output = 'ggplot')
a2 <- getSpatialMap(tgridData, method = 'summer', output = 'ggplot')
a3 <- getSpatialMap(tgridData, method = 'meanAnnual', output = 'ggplot')
a4 <- getSpatialMap(tgridData, method = 'winter', output = 'ggplot')

```

```{r, fig.show='hold', fig.height=6, fig.width=7}

getSpatialMap_comb(a1, a2, a3, a4, nrow = 2)# you cannot assign title
getSpatialMap_comb(a1, a2, a3, a4, nrow = 4)

```

`getSpatialMap_comb` accepts list (using `list =`) object too, which is easier for multi-plot. First list of 12 months are got.
```{r, fig.show='hide'}
c <- lapply(1:12, function(x) getSpatialMap(tgridData, method = x, output = 'ggplot') )
```

Then they are combined.
```{r, fig.show='hold', fig.height=6, fig.width=7}
getSpatialMap_comb(list = c, nrow = 4)
```

### 2.4.2 Bar Plot

Basically, bar plot follows the same rule as part 2.4.1 spatial map, only a few cases that needs to pay attention.

```{r, fig.show='hide', fig.height=4, fig.width=7, results='hide'}
b1 <- getPreciBar(tgridData, method = 'spring', output = 'ggplot')
b2 <- getPreciBar(tgridData, method = 'summer', output = 'ggplot')
b3 <- getPreciBar(tgridData, method = 'autumn', output = 'ggplot')
b4 <- getPreciBar(tgridData, method = 'winter', output = 'ggplot')
```

```{r, fig.show='hold', fig.height=6, fig.width=7}
getPreciBar_comb(b1, b2, b3, b4, nrow = 2)
```

```{r, fig.show='hide', fig.height=4, fig.width=7, results='hide'}
c <- lapply(1:12, function(x) getPreciBar(tgridData, method = x, output = 'ggplot') )
```

```{r, fig.show='hold', fig.height=6, fig.width=7}
getPreciBar_comb(list = c, nrow = 4)
```

## 2.5 Get ensemble forecast from forecasting dataset

If there are different members existing in the dataset, `hyfo` can extract them and generate a dataframe for the easy input to the model.

```
# There is no example data in this test
a <- getFrcEnsem(dataset)
```
The default output is spatially averaged, if there are more than one cells in the dataset, the mean value of the cells will be calculated. While if you are interested in special cell, you can assign the cell value, for how to assign, please check the details in `?getFrcEnsem`

```
a <- getFrcEnsem(dataset, cell = c(2,3))
```
Also, if no plot is needed, you can cancel the plot by the argument `plot = FALSE`

# 3. Anarbe Case

The functions with anarbe case end with `_anarbe`, all of them are used to collect different available published data in anarbe catchment in Spain. The data comes from two website: [here](http://meteo.navarra.es/estaciones/mapadeestaciones.cfm) and [here](http://www4.gipuzkoa.net/oohh/web/esp/02.asp), there are precipitation or discharge data on those website, and can be downloaded directly. 

Since the available files on those website are arranged by a year or five years, for long term data collection, a tools is necessary for collecting data from different files.

##### Note:
For excel files, if you have access to the dam regulation excel file of the dam anarbe, you can use  `collectData_excel_anarbe` in the package, but this function is commented in the original code, cannot be used directly. Go to original file in the library or go to github [here](https://github.com/Yuanchao-Xu/hyfo/blob/master/R/collectData_excel.R), copy the original code.

There are two csv files and txt files included in the package, which can be used as examples.

```{r}
file <- system.file("extdata", "1999.csv", package = "hyfo")
folder <- strsplit(file, '1999')[[1]][1]

a <- collectData_csv_anarbe(folder, output = TRUE)
str(a)
b <- collectData_txt_anarbe(folder, output = TRUE)
str(b)
```



